sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-15 doc/paper-jss/owls.Rmd
Reduce overall file size

```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="90%"}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none") +
  labs(colour = "Estimated effect in log space") +
  # ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
  theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
```

```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="90%",fig.asp=0.3, fig.cap="Map of combined main and spatially varying effects of NAO on Snowy Owl count with confidence intervals. In the west, the lower values span zero implying a no evidence of an effect at this confidence level, while in the southeast the lowest estimates are still positive. Note that the values are in log space and thus additive. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='ht', fig.align='center'}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

```{r plot-all-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='ht', fig.align='center'}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="5in", dpi=200, fig.cap="Spatial random effects that apply across all years.", fig.pos='ht', fig.align='center'}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Spatiotemporal random effects.", fig.pos='ht', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

Finally, the non-random field based elements in the model are represented by `est_non_rf` variable in the prediction dataframe, and in this case only include the overall effect of mean annual NAO and the random intercepts for year (Fig.\ \ref{fig:plot-fixed-effects-owls}).

```{r plot-fixed-effects-owls, out.width="95%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=300, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

```{r plot-map-owls, echo=FALSE, eval=TRUE, out.width="65%", dpi=140, dev="png"}
ggplot(data = p_proj) +
  geom_point(
    data = p_mean, aes(X, Y, colour = nao_effect, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none") +
  labs(colour = "Estimated effect in log space") +
  # ggtitle("Map of combined main and spatially varying effects of NAO on Snowy Owl count") +
  theme(legend.position = c(0.1, 0.07), axis.title = element_blank())
```

```{r plot-map-ci, echo=FALSE, eval=TRUE, out.width="80%",fig.asp=0.3, fig.cap="Map of combined main and spatially varying effects of NAO on Snowy Owl count with confidence intervals. In the west, the lower values span zero implying a no evidence of an effect at this confidence level, while in the southeast the lowest estimates are still positive. Note that the values are in log space and thus additive. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).", fig.pos='ht', fig.align='center', dpi=140, dev="png"}
p1 <- ggplot(data = p_proj) +
  geom_point(
    data = p_mean,
    aes(X, Y, colour = nao_lwr, size = mean_est_count), alpha = 0.5
  ) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000), # adjusts space on sides
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(p_mean$nao_lwr), max(p_mean$nao_upr)), 
    guide = guide_colourbar(direction = "horizontal", title.position = "top")
  ) +
  guides(size = "none", colour = "none") +
  # labs(colour = "Lower 90% CI") +
  ggtitle("Lower 80% CI") +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())

@@

```{r plot-all-effects-owls, out.width="90%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=140, fig.cap="Predicted winter owl counts that incorporate all fixed effects and random effects.", fig.pos='ht', fig.align='center', dev="png"}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

```{r plot-spatial-effects-owls, echo=FALSE, eval=TRUE, out.width="3in", dpi=130, fig.cap="Spatial random effects that apply across all years.", fig.pos='ht', fig.align='center', dev = "png",}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X, Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(
    direction = "horizontal", title.position = "top"
  )) +
  theme(legend.position = c(0.1, 0.1), axis.title = element_blank())
```

@@

```{r plot-spatiotemporal-effects-owls, out.width="75%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=120, fig.cap="Spatiotemporal random effects.", fig.pos='ht', fig.align='center', dev="png"}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = epsilon_st),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

Finally, the non-random field based elements in the model are represented by `est_non_rf` variable in the prediction data frame, and in this case only include the overall effect of mean annual NAO and the random intercepts for year (Fig.\ \ref{fig:plot-fixed-effects-owls}).

```{r plot-fixed-effects-owls, out.width="75%", fig.asp=0.8, echo=FALSE, eval=TRUE, dpi=120, fig.cap="Predictions for each year based on only the non-random field elements.", fig.pos='ht', fig.align='center', dev="png"}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est_non_rf)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    breaks = c(0.03, 0.05, 0.07),
    guide = guide_colourbar(title.position = "top")
  ) +
  theme(axis.text = element_blank(), 
  legend.title = element_text(size = 8, hjust = 0), legend.key.height = unit(0.25, "cm"), 
  axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

### 2022-11-15 doc/paper-jss/paper-jss.Rmd
Reduce overall file size

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

```{r child="model-description.Rmd", eval=TRUE}
```

>>>>

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}; see the model description vignette in `browseVignettes("sdmTMB")` or [rendered on the web](https://pbs-assess.github.io/sdmTMB/articles/)).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.

@@

```{r child="model-description.Rmd", eval=FALSE}
```
